**Best practices for applying machine learning to bacterial 16S rRNA gene sequencing data**

### 2019-08-26 submission/manuscript.Rmd
fixed figures

Second, we trained seven different models using the training data. We selected models with different classification algorithms and regularization methods. Regularization is a technique that discourages overfitting by penalizing the model for learning the training data too well. For regularized logistic regression and SVM with linear kernel, we used L2 regularization to keep all potentially important features. For comparison, we also trained an L1 regularized SVM model with linear kernel. L1-regularization on microbiome data led to a sparser solution (i.e., force many coefficients to zero). To explore the potential for non-linear relationships among features to improve classification, we trained tree-based models including decision tree, random forest, and XGBoost and we trained an SVM model with non-linear kernel.

@@

To show the effect of sample size on model generalizability, we compared cross-validation AUROC values of L2-regularized logistic regression and random forest models when we subsetted our original study design with 490 subjects to 15, 30, 60, 120, and 245 subjects [Figure S4]. The range among the cross-validation AUROC values within both models at lower sample sizes were much larger than when the full collection of samples was used to train and validate the models. These results showed that because the microbiome data had many features (6920 OTUs), it was important to train the models using appropriate sample sizes to avoid problems with generalizability and over-optimism. Furthermore, it was encouraging that even for a small number of samples, the 95% confidence interval included the median AUROC values for the larger subsetted datasets.

@@

To further highlight the differences between the two interpretation methods, we used permutation importance to interpret the linear models [Figure S3]. When we analyzed the L1-regularized SVM with linear kernel model using feature rankings based on weights [Figure 3] and permutation importance [Figure S3], 17 out of the 20 top OTUs (e.g. OTU 609, 822, 1239) were deemed important by both approaches. Similarly, for the L2-regularized SVM and L2-regularized logistic regression, 9 and 12 OTUs, respectively, were shared among the two approaches. These results indicate that both approaches are consistent in selecting the most important OTUs.

>>>>

Second, we trained seven different models using the training data [Table 1]. We selected models with different classification algorithms and regularization methods. Regularization is a technique that discourages overfitting by penalizing the model for learning the training data too well. For regularized logistic regression and SVM with linear kernel, we used L2 regularization to keep all potentially important features. For comparison, we also trained an L1 regularized SVM model with linear kernel. L1-regularization on microbiome data led to a sparser solution (i.e., force many coefficients to zero). To explore the potential for non-linear relationships among features to improve classification, we trained tree-based models including decision tree, random forest, and XGBoost and we trained an SVM model with non-linear kernel.

@@

To show the effect of sample size on model generalizability, we compared cross-validation AUROC values of L2-regularized logistic regression and random forest models when we subsetted our original study design with 490 subjects to 15, 30, 60, 120, and 245 subjects [Figure S3]. The range among the cross-validation AUROC values within both models at lower sample sizes were much larger than when the full collection of samples was used to train and validate the models. These results showed that because the microbiome data had many features (6920 OTUs), it was important to train the models using appropriate sample sizes to avoid problems with generalizability and over-optimism. Furthermore, it was encouraging that even for a small number of samples, the 95% confidence interval included the median AUROC values for the larger subsetted datasets.

@@

To further highlight the differences between the two interpretation methods, we used permutation importance to interpret the linear models [Figure S4]. When we analyzed the L1-regularized SVM with linear kernel model using feature rankings based on weights [Figure 3] and permutation importance [Figure S4], 17 out of the 20 top OTUs (e.g. OTU 609, 822, 1239) were deemed important by both approaches. Similarly, for the L2-regularized SVM and L2-regularized logistic regression, 9 and 12 OTUs, respectively, were shared among the two approaches. These results indicate that both approaches are consistent in selecting the most important OTUs.
